Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-08-22 paper/ch_potentials.tex
typos and wordsmithing and labels from diss

\paragraph{Speedup Techniques for Shortest Paths in Road Networks}
A*~\cite{hnr-afbhd-68} is a well-known approach to accelerate shortest path queries by directing the search towards the target through heuristic distance estimates.
It has been successfully employed in many problems beyond route planning in transportation networks~\cite{DBLP:conf/socs/StrasserHB14,DBLP:conf/ijcai/BonoGHS19,DBLP:conf/ijcai/0002UJAKK18}.
It is one of the algorithms fundamental to our work.
The performance of A* crucially depends on the quality of the heuristic estimates.
For example, while geographic distances may seem like a natural choice for distance estimates in road networks, the resulting heuristic is rather ineffective.
In fact, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
ALT was one of the early speedup technique for routing in road networks.
It uses precomputed distances to certain landmark vertices combined with the triangle inequality to compute distance estimates.
On road networks, this achieves speedups of around two orders of magnitude over Dijkstra's algorithm.
However, the speedups obtained by ALT are still limited by the quality of the distance estimates.
To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal directed search on.
This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
Since then, there has been little development on A*-based techniques for routing in road networks.
This is because hierarchical techniques have proven much more effective.

Hierarchical speedup techniques utilize the inherent hierarchy in road networks, usually in combination with the \emph{contraction} of less important vertices.
One popular example is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
The ``importance'' of vertices is determined heuristically during preprocessing.
This results in queries taking less than a tenth of a millisecond which is a speedup of more than four orders of magnitude over Dijkstra's algorithm.
The preprocessing takes only a few minutes and produces little more auxiliary data than the input graph.
CH has been used successfully in many real world applications and quite a few open-source implementations exist\footnote{
\url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
}.
CH is another fundamental algorithm for our work.
More specifically, we extend on PHAST~\cite{dgnw-phast-13} and RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
These algorithms reuse the CH preprocessing but modify the query phase.
PHAST supports \emph{one-to-all} queries from one source vertex to \emph{all} other vertices.
RPHAST is a variant for \emph{one-to-many} queries to a subset of vertices known in advance.
%
Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02}.
It is based on a multi-level partition of the road network computed during preprocessing.
Queries are slightly slower than CH queries but still three to four orders of magnitude faster than Dijkstra's algorithm.
The memory consumption is even smaller than with CH.
Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, these works also primarily focused on obtaining faster queries rather than simple and flexible approaches.

@@

\node (s) at (0,0)     [node] {$s$};
\node (2) at (1.5,1.5) [node] {};
\node (3) at (3.7,1)   [node] {};
\node (4) at (5,3)     [node] {$v^*$};
\node (5) at (5.75,-0.1)  [node] {};
\node (6) at (6.75,1.2)   [node] {};
\node (7) at (7.75,-0.5)    [node] {};
\node (8) at (8.5,2)   [node] {};
\node (t) at (10,0.5)  [node] {$t$};

@@

\Fn{\Dist{$u$}}{
	\If{$D[u] = \bot$}{
		$D[u]\leftarrow D^{\downarrow}[u]$\;
		\For{all up-edges $uv \in \echu$}{
      $D[u]\leftarrow\min(D[u],w(uv)+\Dist(v))$\;
		}
	}
	\Return{$D[u]$}\;
}
\caption{Lazy RPHAST algorithm.}
\label{algo:pot}
\end{algorithm2e}

@@

Now, distances from many sources $s_i$ to $t$ can be computed incrementally with the \textsf{ComputeAndMemoizeDist} function as shown in Algorithm~\ref{algo:pot}.
The key to doing this efficiently is reusing the distance information $D$ across invocations through memoization.
Thus, the first step of the algorithm is always to check if the distance for the requested vertex has already been computed and immediately returning it if this is the case.
If not, the distance $D[s_i]$ is initialized to the shortest down-path distance $D^{\downarrow}[s_i]$ obtained by the backward search.
Then, the algorithm iterates over all up-edges $(s_i v)$ and checks if the up-down path through this neighbor can improve the distance.
To obtain the shortest distance from a neighbor $v$ to $t$, the algorithm is invoked recursively.

@@

Algorithm~\ref{algo:pot} can be applied to CCH without any modifications.
However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
To compute the distance $D[v]$ of a vertex $v$, the distances of all upward neighbors must be final.
In Algorithm~\ref{algo:pot}, these upward neighbor distances are computed recursively.
Thus, the search space is explored in a DFS-like fashion and distances are finalized in DFS post order.
However, the path from a vertex to the root in the elimination tree is a also topological order for the upward search space.
This is because the ancestors in the elimination tree contain the entire upward search space~\cite{bcrw-s-16}.
Therefore, iterating over the vertices on the elimination tree path from the root to $v$ while relaxing outgoing upward edges of each vertex also yields shortest distances.
Further, with this approach, when a vertex $v$ already has a distance $D[v] \neq \bot$, all ancestors of $v$ must already have their final distance, too.
Thus, as soon as the algorithm encounters a vertex which already has a final distance, the remaining search space is known to have final distances.

@@

% The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:pot}.
% We describe vectorized variants of these algorithms.
% We replace the $B$ and $D$ arrays of scalars by arrays of $k$-dimensional vectors of scalars.
% To clarify which entities are SIMD vectors, we denote the vectorized arrays by $\hat{B}$ and $\hat{D}$.
% $\hat{D}[s]$ is an SIMD vector $\hat{v}$ of length $k$.
% The $i$-th element $\hat{v}[i]$ of $\hat{v}$ will contain the distance from $s$ to $i$-th selected target.
% We can also denote this distance using $\hat{D}[s][i]$.
% All operations on SIMD vectors are component-wise.
% Adding a scalar to an SIMD vector consists of adding the scalar to every element of the SIMD vector.

@@

% Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
% In the original algorithm, the vertices are ordered by distance.
% In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
% To resolve this problem, we ordered the vertices in the queue not by distance but by CH rank.
% The CH rank is the position of a vertex in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
% All other modifications are straightforward.
% The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
% Algorithm~\ref{algo:pot} can be vectorized by just replacing the arrays by their SIMD variants.
% The result is illustrated in Algorithm~\ref{algo:simd-pot}.

@@

% Luckily, we can do better by integrating a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
% The Bucket CH selection phase works as follows:
% % TODO better structure or local outline
% % TODO besser herleiten, wieso kein 2d array
% For each vertex $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$.
% These pairs are called \emph{buckets}.
% These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
% Now Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
% We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
% After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target vertex.
% This way, the search space of the sources is explored only once.

@@

\subsection{CH-Potentials}\label{sec:chpot}
CH-Potentials can be used to solve any problem in this setup.
The preprocessing is always the computation of the CH augmented graph $G^+_{\ell}$ and remains the same regardless of the specific routing problem.
The query consists of A* with the heuristic function $h_t(v) = \dist_{w_{\ell}}(\phi(v), \phi(t))$ computed using Lazy RPHAST.
In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $\phi(v)$ (see Algorithm~\ref{algo:pot}).
Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
As the performance of A* depends crucially on the accuracy of the heuristic estimates, the smaller the difference between query weights and lower bound distances, the better CH-Potentials will perform.

@@

The performance of A* depends on the tightness of the heuristic and the overhead of evaluating the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
For $\alpha > 1.1$, the growth slows down.
This illustrates both the strengths and limits of our approach and goal directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
The factor 1.05 was chosen to resemble realistic problem settings where goal directed search can achieve reasonable speedups (compare to Table~\ref{tab:applications}).
We compare (C)CH-Potentials to three other heuristics.
The first heuristic is the \emph{Zero} heuristic where $h(v)$ is $0$ for all vertices $v$.
This corresponds to using Dijkstra's algorithm.
Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target vertex but do not include the running time for this step.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
(C)CH-Potentials computes the same distance estimates, but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

>>>>

\paragraph{Speedup Techniques for Shortest Paths in Road Networks}
A*~\cite{hnr-afbhd-68} is a well-known approach to accelerate shortest path queries by directing the search towards the target through heuristic distance estimates.
It has been successfully employed in many problems beyond route planning in transportation networks~\cite{DBLP:conf/socs/StrasserHB14,DBLP:conf/ijcai/BonoGHS19,DBLP:conf/ijcai/0002UJAKK18}.
It is one of the algorithms fundamental to our work.
The performance of A* crucially depends on the quality of the heuristic estimates.
For example, while geographic distances may seem like a natural choice for distance estimates in road networks, the resulting heuristic is rather ineffective.
In fact, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
ALT was one of the early speedup technique for routing in road networks.
It uses precomputed distances to certain landmark vertices combined with the triangle inequality to compute distance estimates.
On road networks, this achieves speedups of around two orders of magnitude over Dijkstra's algorithm.
However, the speedups obtained by ALT are still limited by the quality of the distance estimates.
To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal-directed search on.
This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
Since then, there has been little development on A*-based techniques for routing in road networks.
This is because hierarchical techniques have proven more effective.

Hierarchical speedup techniques utilize the inherent hierarchy in road networks, usually in combination with the \emph{contraction} of less important vertices.
One popular example is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
The ``importance'' of vertices is determined heuristically during preprocessing.
This results in queries taking less than a tenth of a millisecond which is a speedup of more than four orders of magnitude over Dijkstra's algorithm.
The preprocessing takes only a few minutes and produces little more auxiliary data than the input graph.
CH has been used successfully in many real world applications and quite a few open-source implementations exist.\footnote{
\url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
}
CH is another fundamental algorithm for our work.
More specifically, we extend on PHAST~\cite{dgnw-phast-13} and RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
These algorithms reuse the CH preprocessing but modify the query phase.
PHAST supports \emph{one-to-all} queries from one source vertex to \emph{all} other vertices.
RPHAST is a variant for \emph{one-to-many} queries to a subset of vertices known in advance.
%
Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02}.
It is based on a multi-level partition of the road network computed during preprocessing.
Queries are slightly slower than CH queries but still three to four orders of magnitude faster than Dijkstra's algorithm.
The memory consumption is even smaller than with CH.
Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, these works also primarily focused on obtaining faster queries rather than simple and flexible approaches.

@@

\node (s) at (0,0)     [node] {$s$};
\node (2) at (1.5,1.5) [node] {};
\node (3) at (3.7,1)   [node] {};
\node (4) at (5,3)     [node] {};
\node (5) at (5.75,-0.1)  [node] {};
\node (6) at (6.75,1.2)   [node] {};
\node (7) at (7.75,-0.5)    [node] {};
\node (8) at (8.5,2)   [node] {};
\node (t) at (10,0.5)  [node] {$t$};

@@

\Fn{\Dist{$u$}}{
	\If{$D[u] = \bot$}{
		$D[u]\leftarrow D^{\downarrow}[u]$\;
		\For{all up-edges $uv \in \echu$}{
      $D[u]\leftarrow\min(D[u],w(uv)+\Dist(v))$\;
		}
	}
	\Return{$D[u]$}\;
}
\caption{Lazy RPHAST algorithm.}
\label{algo:lazy_rphast_ch}
\end{algorithm2e}

@@

Now, distances from many sources $s_i$ to $t$ can be computed incrementally with the \textsf{ComputeAndMemoizeDist} function as shown in Algorithm~\ref{algo:lazy_rphast_ch}.
The key to doing this efficiently is reusing the distance information $D$ across invocations through memoization.
Thus, the first step of the algorithm is always to check if the distance for the requested vertex has already been computed and immediately returning it if this is the case.
If not, the distance $D[s_i]$ is initialized to the shortest down-path distance $D^{\downarrow}[s_i]$ obtained by the backward search.
Then, the algorithm iterates over all up-edges $(s_i v)$ and checks if the up-down path through this neighbor can improve the distance.
To obtain the shortest distance from a neighbor $v$ to $t$, the algorithm is invoked recursively.

@@

Algorithm~\ref{algo:lazy_rphast_ch} can be applied to CCH without any modifications.
However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
To compute the distance $D[v]$ of a vertex $v$, the distances of all upward neighbors must be final.
In Algorithm~\ref{algo:lazy_rphast_ch}, these upward neighbor distances are computed recursively.
Thus, the search space is explored in a DFS-like fashion and distances are finalized in DFS post order.
However, the path from a vertex to the root in the elimination tree is a also topological order for the upward search space.
This is because the ancestors in the elimination tree contain the entire upward search space~\cite{bcrw-s-16}.
Therefore, iterating over the vertices on the elimination tree path from the root to $v$ while relaxing outgoing upward edges of each vertex also yields shortest distances.
Further, with this approach, when a vertex $v$ already has a distance $D[v] \neq \bot$, all ancestors of $v$ must already have their final distance, too.
Thus, as soon as the algorithm encounters a vertex which already has a final distance, the remaining search space is known to have final distances.

@@

% The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:lazy_rphast_ch}.
% We describe vectorized variants of these algorithms.
% We replace the $B$ and $D$ arrays of scalars by arrays of $k$-dimensional vectors of scalars.
% To clarify which entities are SIMD vectors, we denote the vectorized arrays by $\hat{B}$ and $\hat{D}$.
% $\hat{D}[s]$ is an SIMD vector $\hat{v}$ of length $k$.
% The $i$-th element $\hat{v}[i]$ of $\hat{v}$ will contain the distance from $s$ to $i$-th selected target.
% We can also denote this distance using $\hat{D}[s][i]$.
% All operations on SIMD vectors are component-wise.
% Adding a scalar to an SIMD vector consists of adding the scalar to every element of the SIMD vector.

@@

% Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
% In the original algorithm, the vertices are ordered by distance.
% In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
% To resolve this problem, we ordered the vertices in the queue not by distance but by CH rank.
% The CH rank is the position of a vertex in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
% All other modifications are straightforward.
% The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
% Algorithm~\ref{algo:lazy_rphast_ch} can be vectorized by just replacing the arrays by their SIMD variants.
% The result is illustrated in Algorithm~\ref{algo:simd-pot}.

@@

% Luckily, we can do better by integrating a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
% The Bucket CH selection phase works as follows:
% % TODO better structure or local outline
% % TODO besser herleiten, wieso kein 2d array
% For each vertex $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$.
% These pairs are called \emph{buckets}.
% These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
% Now Algorithm~\ref{algo:lazy_rphast_ch} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
% We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
% After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target vertex.
% This way, the search space of the sources is explored only once.

@@

\subsection{CH-Potentials}\label{sec:chpot}
CH-Potentials can be used to solve any problem in this setup.
The preprocessing is always the computation of the CH augmented graph $G^+_{\ell}$ and remains the same regardless of the specific routing problem.
The query consists of A* with the heuristic function $h_t(v) = \dist_{w_{\ell}}(\phi(v), \phi(t))$ computed using Lazy RPHAST.
In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $\phi(v)$ (see Algorithm~\ref{algo:lazy_rphast_ch}).
Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
As the performance of A* depends crucially on the accuracy of the heuristic estimates, the smaller the difference between query weights and lower bound distances, the better CH-Potentials will perform.

@@

The performance of A* depends on the tightness of the heuristic and the overhead of evaluating the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
For $\alpha > 1.1$, the growth slows down.
This illustrates both the strengths and limits of our approach and goal-directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
The factor 1.05 was chosen to resemble realistic problem settings where goal-directed search can achieve reasonable speedups (compare to Table~\ref{tab:applications}).
We compare (C)CH-Potentials to three other heuristics.
The first heuristic is the \emph{Zero} heuristic where $h(v)$ is $0$ for all vertices $v$.
This corresponds to using Dijkstra's algorithm.
Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target vertex but do not include the running time for this step.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
(C)CH-Potentials computes the same distance estimates, but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.
